---
title: "Introduction to SingleArmMRCT"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to SingleArmMRCT}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
library(SingleArmMRCT)
```

## Background

Multi-regional clinical trials (MRCTs) are increasingly used in global drug development to allow simultaneous regulatory submissions across multiple regions. A key requirement for regional approval — particularly in Japan under the MHLW guidelines — is the demonstration of **regional consistency**: evidence that the treatment effect observed in a specific region (e.g., Japan) is consistent with the overall trial result.

Two widely used consistency evaluation methods are:

- **Method 1** (Effect Retention): Evaluates whether Region 1 retains at least a fraction $\pi$ of the overall treatment effect.
- **Method 2** (Simultaneous Positivity): Evaluates whether all regional estimates simultaneously exceed the null value.

These methods were originally developed for **two-arm randomised trials**. However, single-arm trials are now common in oncology and rare disease settings, where randomisation is infeasible or unethical. The **SingleArmMRCT** package extends Method 1 and Method 2 to the single-arm setting, where the treatment effect is defined relative to a historical control value.

---

## Regional Consistency Probability

The **Regional Consistency Probability (RCP)** is defined as the probability that a consistency criterion is satisfied, computed under the true parameter values assumed at the design stage.

### Method 1

For a given endpoint with treatment effect parameter $\theta$ (e.g., mean, response rate, hazard ratio), Method 1 requires that Region 1 retains at least a fraction $\pi$ of the overall effect:

$$
\text{RCP}_1 = \Pr\!\left[(\hat{\theta}_1 - \theta_0) > \pi \times (\hat{\theta} - \theta_0)\right]
$$

where $\hat{\theta}_1$ is the regional estimate for Region 1, $\hat{\theta}$ is the overall estimate, and $\theta_0$ is the null (historical control) value. The threshold $\pi$ is typically set to $0.5$.

### Method 2

Method 2 requires that all $J$ regional estimates simultaneously show a positive effect:

$$
\text{RCP}_2 = \Pr\!\left[\hat{\theta}_j \succ \theta_0 \text{ for all } j = 1, \ldots, J\right]
$$

where $\succ$ denotes the direction of benefit (e.g., $>$ for continuous endpoints, $<$ for hazard ratios).

---

## Package Structure

The package provides a pair of functions for each of six endpoint types.

```{r echo=FALSE}
tbl <- data.frame(
  Endpoint = c(
    "Continuous",
    "Binary",
    "Count (negative binomial)",
    "Time-to-event (hazard ratio)",
    "Milestone survival",
    "Restricted mean survival time (RMST)"
  ),
  `Calculation function` = c(
    "rcp1armContinuous()",
    "rcp1armBinary()",
    "rcp1armCount()",
    "rcp1armHazardRatio()",
    "rcp1armMilestoneSurvival()",
    "rcp1armRMST()"
  ),
  `Plot function` = c(
    "plot_rcp1armContinuous()",
    "plot_rcp1armBinary()",
    "plot_rcp1armCount()",
    "plot_rcp1armHazardRatio()",
    "plot_rcp1armMilestoneSurvival()",
    "plot_rcp1armRMST()"
  ),
  check.names = FALSE
)
knitr::kable(tbl)
```

Each calculation function supports two approaches:

- **`"formula"`**: Closed-form or semi-analytical solution based on normal approximation.
- **`"simulation"`**: Monte Carlo simulation.

---

## Common Parameters

All functions share the following key parameters.

```{r echo=FALSE}
params <- data.frame(
  Parameter = c("Nj", "PI", "approach", "nsim", "seed"),
  Description = c(
    "Integer vector of regional sample sizes, e.g. c(10, 90) for J = 2 regions",
    "Effect retention threshold pi for Method 1 (default 0.5)",
    'Calculation approach: "formula" or "simulation"',
    "Number of Monte Carlo iterations (simulation only; default 10000)",
    "Random seed for reproducibility (simulation only; default 1)"
  )
)
knitr::kable(params)
```

Time-to-event endpoints additionally require:

```{r echo=FALSE}
params_tte <- data.frame(
  Parameter = c("t_a", "t_f", "lambda_dropout"),
  Description = c(
    "Accrual period (duration of patient enrollment)",
    "Follow-up period (additional follow-up after accrual ends)",
    "Dropout hazard rate; NULL = no dropout (default NULL)"
  )
)
knitr::kable(params_tte)
```

---

## Quick Example

The following example computes RCP for a **continuous endpoint** with $N = 100$ subjects, Region 1 allocated $f_1 = 10\%$, true mean $\mu = 0.5$, null mean $\mu_0 = 0.1$, and standard deviation $\sigma = 1$.

```{r}
result <- rcp1armContinuous(
  mu       = 0.5,
  mu0      = 0.1,
  sd       = 1,
  Nj       = c(10, 90),
  PI       = 0.5,
  approach = "formula"
)
print(result)
```

The formula and simulation results can be compared directly:

```{r}
result_sim <- rcp1armContinuous(
  mu       = 0.5,
  mu0      = 0.1,
  sd       = 1,
  Nj       = c(10, 90),
  PI       = 0.5,
  approach = "simulation",
  nsim     = 10000,
  seed     = 1
)
print(result_sim)
```

---

## Visualisation

Each endpoint has a corresponding plot function that displays RCP as a function of the regional allocation proportion $f_1 = N_1 / N$, with facets for different total sample sizes $N$.

```{r fig.width=9, fig.height=4}
plot_rcp1armContinuous(
  mu    = 0.5,
  mu0   = 0.1,
  sd    = 1,
  PI    = 0.5,
  N_vec = c(20, 40, 100),
  J     = 3,
  nsim  = 5000,
  seed  = 1
)
```

Solid lines represent the closed-form formula; dashed lines represent Monte Carlo simulation. Blue lines correspond to Method 1 and yellow lines to Method 2. The horizontal grey dashed line marks RCP = 0.80, a commonly used design target.

---

## Further Reading

For endpoint-specific details, see the companion vignettes:

- **Non-survival endpoints**: continuous, binary, and count endpoints.
- **Survival endpoints**: hazard ratio, milestone survival, and RMST endpoints.

### References

Hayashi R, Itoh Y (2018). A reexamination of Japanese sample size calculation for multiregional clinical trial evaluating survival endpoint. *Pharmaceutical Statistics*, 17(1): 46--55.

Teng Z, Bhatt DL, Rothmann M, Mehta C (2018). Practical recommendations for regional consistency evaluation in multi-regional clinical trials with time-to-event endpoints. *Journal of Biopharmaceutical Statistics*, 28(6): 1089--1106.

Tang Y (2022). Complex survival trial design by the product integration method. *Statistics in Medicine*, 41(4): 798--814.

Wu J (2015). Sample size calculation for the one-sample log-rank test. *Pharmaceutical Statistics*, 14(1): 26--33.
