---
title: "Survival Endpoints: Hazard Ratio, Milestone Survival, and RMST"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Survival Endpoints: Hazard Ratio, Milestone Survival, and RMST}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
library(SingleArmMRCT)
```

This vignette describes Regional Consistency Probability (RCP) calculations for three survival endpoint types: **hazard ratio**, **milestone survival**, and **restricted mean survival time (RMST)**. All three endpoints share a common trial design framework with accrual and follow-up periods.

---

## Common trial design parameters

All survival endpoint functions require the following design parameters in addition to the endpoint-specific inputs.

| Parameter | Description |
|---|---|
| `t_a` | Accrual period: duration over which patients are enrolled (uniform accrual assumed) |
| `t_f` | Follow-up period: additional follow-up after accrual ends |
| `tau` | Total study duration: $\tau = t_a + t_f$ (computed internally) |
| `lambda_dropout` | Exponential dropout hazard rate; `NULL` = no dropout (default) |

Event times are modelled by the **exponential distribution** with hazard rate $\lambda$ under the alternative hypothesis. The historical control hazard rate is $\lambda_0$, so the true hazard ratio is $HR = \lambda / \lambda_0 < 1$ (benefit assumed).

In the examples below, a common design is used:

```{r}
lambda  <- log(2) / 10   # treatment median survival = 10
lambda0 <- log(2) / 5    # historical control median survival = 5
t_a     <- 3             # accrual period
t_f     <- 10            # follow-up period
```

---

## 1. Hazard Ratio Endpoint

### Statistical model

Under the exponential model, the log-hazard ratio estimator for Region $j$ is approximately:

$$
\log(\widehat{HR}_j) \sim N\!\left(\log(HR),\; \frac{1}{E_j}\right)
$$

where $E_j = N_j \phi$ is the expected number of events in Region $j$, and $\phi$ is the event probability per patient (Wu 2015):

$$
\phi = \frac{\lambda}{\lambda + \lambda_d}
\left[1 - \frac{e^{-(\lambda + \lambda_d)t_f} - e^{-(\lambda + \lambda_d)\tau}}{(\lambda + \lambda_d)\,t_a}\right]
$$

with $\lambda_d$ denoting the dropout hazard rate (0 if no dropout).

### Consistency criteria

Method 1 is evaluated on two scales:

**Method 1 (log-HR scale)**:

$$
\text{RCP}_{1,\log} = \Pr\!\left[\log(\widehat{HR}_1) < \pi \times \log(\widehat{HR})\right]
= \Phi\!\left(\frac{-(1-\pi)\log(HR)}{\sqrt{(1-\pi f_1)^2/E_1 + \{\pi(1-f_1)\}^2/E_2}}\right)
$$

**Method 1 (linear-HR scale)** (Hayashi and Itoh 2018):

$$
\text{RCP}_{1,\text{lin}} = \Pr\!\left[(1 - \widehat{HR}_1) > \pi \times (1 - \widehat{HR})\right]
$$

This is derived via the delta method applied to the transformation $g = \log(\widehat{HR}_1) - \log(1 - \pi + \pi\,\widehat{HR})$.

**Method 2**:

$$
\text{RCP}_2 = \prod_{j=1}^{J} \Phi\!\left(-\log(HR)\,\sqrt{E_j}\right)
$$

### Example

```{r}
result_f <- rcp1armHazardRatio(
  lambda         = lambda,
  lambda0        = lambda0,
  Nj             = c(20, 80),
  t_a            = t_a,
  t_f            = t_f,
  lambda_dropout = NULL,
  PI             = 0.5,
  approach       = "formula"
)
print(result_f)
```

```{r}
result_s <- rcp1armHazardRatio(
  lambda         = lambda,
  lambda0        = lambda0,
  Nj             = c(20, 80),
  t_a            = t_a,
  t_f            = t_f,
  lambda_dropout = NULL,
  PI             = 0.5,
  approach       = "simulation",
  nsim           = 10000,
  seed           = 1
)
print(result_s)
```

### Effect of dropout

Specifying `lambda_dropout` reduces the effective event probability $\phi$, leading to lower RCP values.

```{r}
result_dropout <- rcp1armHazardRatio(
  lambda         = lambda,
  lambda0        = lambda0,
  Nj             = c(20, 80),
  t_a            = t_a,
  t_f            = t_f,
  lambda_dropout = 0.05,
  PI             = 0.5,
  approach       = "formula"
)
print(result_dropout)
```

### Visualisation

```{r fig.width=9, fig.height=6}
plot_rcp1armHazardRatio(
  lambda  = lambda,
  lambda0 = lambda0,
  t_a     = t_a,
  t_f     = t_f,
  PI      = 0.5,
  N_vec   = c(20, 40, 100),
  J       = 3,
  nsim    = 5000,
  seed    = 1
)
```

---

## 2. Milestone Survival Endpoint

### Statistical model

The treatment effect at a prespecified evaluation time $t_{\text{eval}}$ is expressed as the difference in survival probabilities:

$$
\delta = S(t_{\text{eval}}) - S_0(t_{\text{eval}})
$$

where $S(t) = e^{-\lambda t}$ is the true survival function and $S_0(t_{\text{eval}})$ is the historical control survival rate at $t_{\text{eval}}$.

The Kaplan-Meier estimator $\hat{S}_j(t_{\text{eval}})$ has asymptotic variance derived from Greenwood's formula (Tang 2022):

$$
\mathrm{Var}\!\left[\hat{S}_j(t)\right] \approx
\frac{e^{-2\lambda t}}{N_j}
\int_0^{t} \frac{\lambda\, e^{(\lambda + \lambda_d)u}}{G_a(u)}\,du
$$

where $G_a(u) = 1$ for $u \leq t_f$ and $G_a(u) = (\tau - u)/t_a$ for $u > t_f$.

When $t_{\text{eval}} \leq t_f$, this integral has a **closed-form solution**:

$$
\mathrm{Var}\!\left[\hat{S}_j(t)\right] =
\frac{e^{-2\lambda t}}{N_j} \cdot \frac{\lambda}{\lambda + \lambda_d}
\left(e^{(\lambda + \lambda_d)t} - 1\right)
$$

When $t_{\text{eval}} > t_f$, numerical integration is used.

### Consistency criteria

**Method 1**:

$$
\text{RCP}_1 = \Pr\!\left[(\hat{S}_1(t) - S_0(t)) > \pi \times (\hat{S}(t) - S_0(t))\right]
= \Phi\!\left(\frac{(1-\pi)\,\delta}{\sqrt{(1-\pi f_1)^2 V_1 + \{\pi(1-f_1)\}^2 V_2}}\right)
$$

where $V_j = \mathrm{Var}[\hat{S}_j(t_{\text{eval}})]$.

**Method 2**:

$$
\text{RCP}_2 = \prod_{j=1}^{J} \Phi\!\left(\frac{\delta}{\sqrt{V_j}}\right)
$$

### Example

```{r}
t_eval <- 8
S0     <- exp(-log(2) * t_eval / 5)   # historical control S(8) with median = 5

result_f <- rcp1armMilestoneSurvival(
  lambda         = lambda,
  t_eval         = t_eval,
  S0             = S0,
  Nj             = c(20, 80),
  t_a            = t_a,
  t_f            = t_f,
  lambda_dropout = NULL,
  PI             = 0.5,
  approach       = "formula"
)
print(result_f)
```

```{r}
result_s <- rcp1armMilestoneSurvival(
  lambda         = lambda,
  t_eval         = t_eval,
  S0             = S0,
  Nj             = c(20, 80),
  t_a            = t_a,
  t_f            = t_f,
  lambda_dropout = NULL,
  PI             = 0.5,
  approach       = "simulation",
  nsim           = 10000,
  seed           = 1
)
print(result_s)
```

The output includes `formula_type`, which indicates whether the closed-form or numerical integration was used.

### Visualisation

```{r fig.width=9, fig.height=4}
plot_rcp1armMilestoneSurvival(
  lambda = lambda,
  t_eval = t_eval,
  S0     = S0,
  t_a    = t_a,
  t_f    = t_f,
  PI     = 0.5,
  N_vec  = c(20, 40, 100),
  J      = 3,
  nsim   = 5000,
  seed   = 1
)
```

---

## 3. RMST Endpoint

### Statistical model

The restricted mean survival time (RMST) up to a truncation time $\tau^*$ is:

$$
\mu(\tau^*) = \int_0^{\tau^*} S(t)\,dt = \frac{1 - e^{-\lambda\tau^*}}{\lambda}
$$

The regional RMST estimator $\hat{\mu}_j(\tau^*)$ is the area under the Kaplan-Meier curve up to $\tau^*$. Its asymptotic variance is (Tang 2022):

$$
\mathrm{Var}\!\left[\hat{\mu}_j(\tau^*)\right] \approx
\frac{1}{N_j}
\int_0^{\tau^*}
\frac{\left(e^{-\lambda t} - e^{-\lambda\tau^*}\right)^2}
     {\lambda\, e^{-\lambda t}\, G(t)}
\,dt
$$

where $G(t) = e^{-\lambda_d t} \cdot G_a(t)$ is the overall censoring survival function.

When $\tau^* \leq t_f$, this has a **closed-form solution**:

$$
v(\tau^*) = \frac{1}{\lambda}\left[
  A(\lambda_d) - 2e^{-\lambda\tau^*}A(\lambda_d + \lambda)
  + e^{-2\lambda\tau^*}A(\lambda_d + 2\lambda)
\right]
$$

where $A(r) = (e^{r\tau^*} - 1)/r$ for $r > 0$ and $A(0) = \tau^*$.

When $\tau^* > t_f$, numerical integration is used.

### Consistency criteria

**Method 1**:

$$
\text{RCP}_1 = \Pr\!\left[(\hat{\mu}_1 - \mu_0) > \pi \times (\hat{\mu} - \mu_0)\right]
= \Phi\!\left(\frac{(1-\pi)(\mu_{\text{est}} - \mu_0)}{\sqrt{(1-\pi f_1)^2 W_1 + \{\pi(1-f_1)\}^2 W_2}}\right)
$$

where $W_j = \mathrm{Var}[\hat{\mu}_j(\tau^*)]$ and $\mu_0 = \mu_0(\tau^*)$ is the historical control RMST.

**Method 2**:

$$
\text{RCP}_2 = \prod_{j=1}^{J} \Phi\!\left(\frac{\mu_{\text{est}} - \mu_0}{\sqrt{W_j}}\right)
$$

### Example

```{r}
tau_star <- 8
mu0      <- (1 - exp(-lambda0 * tau_star)) / lambda0   # historical control RMST

result_f <- rcp1armRMST(
  lambda         = lambda,
  tau_star       = tau_star,
  mu0            = mu0,
  Nj             = c(20, 80),
  t_a            = t_a,
  t_f            = t_f,
  lambda_dropout = NULL,
  PI             = 0.5,
  approach       = "formula"
)
print(result_f)
```

```{r}
result_s <- rcp1armRMST(
  lambda         = lambda,
  tau_star       = tau_star,
  mu0            = mu0,
  Nj             = c(20, 80),
  t_a            = t_a,
  t_f            = t_f,
  lambda_dropout = NULL,
  PI             = 0.5,
  approach       = "simulation",
  nsim           = 10000,
  seed           = 1
)
print(result_s)
```

### Visualisation

```{r fig.width=9, fig.height=4}
plot_rcp1armRMST(
  lambda   = lambda,
  tau_star = tau_star,
  mu0      = mu0,
  t_a      = t_a,
  t_f      = t_f,
  PI       = 0.5,
  N_vec    = c(20, 40, 100),
  J        = 3,
  nsim     = 5000,
  seed     = 1
)
```

---

## Comparison across survival endpoints

```{r echo=FALSE}
tbl <- data.frame(
  Endpoint = c("Hazard Ratio", "Milestone Survival", "RMST"),
  `Effect measure` = c(
    "log(HR) = log(lambda / lambda0)",
    "delta = S(t_eval) - S0(t_eval)",
    "delta = mu(tau*) - mu0(tau*)"
  ),
  `Variance basis` = c(
    "Expected events (Wu 2015)",
    "Greenwood formula (Tang 2022)",
    "Integral of squared survival difference (Tang 2022)"
  ),
  `Closed-form condition` = c(
    "Always (normal approximation)",
    "t_eval <= t_f",
    "tau_star <= t_f"
  ),
  check.names = FALSE
)
knitr::kable(tbl)
```

---

## References

Hayashi R, Itoh Y (2018). A reexamination of Japanese sample size calculation for multiregional clinical trial evaluating survival endpoint. *Pharmaceutical Statistics*, 17(1): 46--55.

Tang Y (2022). Complex survival trial design by the product integration method. *Statistics in Medicine*, 41(4): 798--814.

Wu J (2015). Sample size calculation for the one-sample log-rank test. *Pharmaceutical Statistics*, 14(1): 26--33.
